---
title: "Hypothesis Test"
author: "Yaochieh Yao"
date: "2/21/2023"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
data(golub, package = "multtest")
```
## Module 6 Hypothesis Test<br>
**Problem 1 <br>**
On the Golub et al. (1999) data, consider the “H4/j gene” gene (row
2972) and the “APS Prostate specific antigen” gene (row 2989). Setup the
appropriate hypothesis for proving the following claims. Chose and carry out the
appropriate tests.<br>
<br>
a.The mean “H4/j gene” gene expression value in the ALL group is greater than
-0.9 (note that this is negative 0.9).<br>
<br>
ANS: <br>
H0: The mean “H4/j gene” gene expression value in the ALL group is -0.09<br>
HA: The mean “H4/j gene” gene expression value in the ALL group greater<br>
than -0.09 <br>
<br>
Since the p-value 0.01601 is significant (less than 0.05), we can reject the H0
hypothesis, and conclude that the mean “H4/j gene” gene expression value in the
ALL group is greater than -0.09
```{r}
gol.fac <- factor(golub.cl,levels=0:1, labels= c("ALL","AML"))
# Problem 1-a
t.test(golub[2972,gol.fac=="ALL"],mu=-0.9, alternative = "greater") # p-value = 0.01601
```
<br>
<br>
b.The mean “H4/j gene” gene expression value in ALL group differs from the
mean “H4/j gene” gene expression value in the AML group.<br>
<br>
ANS:<br>
H0: The mean “H4/j gene” gene expression value in ALL group differs from AML
group<br>
HA: The mean “H4/j gene” gene expression value in ALL group is the same as
AML group<br>
<br>
Since the p-value 0.1444 is large, we did not have enough evidence to reject the
null hypothesis.
```{r}
# Problem 1-b
t.test(golub[2972,]~gol.fac) # p-value = 0.1444
```
<br>
<br>
c.In the ALL group, the mean expression value for the “H4/j gene” gene is lower
than the mean expression value for the “APS Prostate specific antigen” gene.<br>
<br>
ANS:<br>
H0: In the ALL group, the mean expression value for the “H4/j gene” gene is equal
to the “APS Prostate specific antigen” gene<br>
HA: In the ALL group, the mean expression value for the “H4/j gene” gene is less
than the “APS Prostate specific antigen” gene<br>
<br>
Since the p-value 0.04541 is significant (less than 0.05), we can reject the null
hypothesis and conclude that the mean expression value for the “H4/j gene” gene is
less than the “APS Prostate specific antigen” gene in the ALL group.
```{r}
# Problem 1-c
t.test(golub[2972, gol.fac=="ALL"],golub[2989,gol.fac=="ALL"],alternative="less") # p-value = 0.04541
```
<br>
<br>
d.Let pH4j denotes the proportion of patients for whom the “H4/j gene” expression
values is greater than -0.6. We wish to show that pH4j in the ALL group is less than
0.5.<br>
<br>
ANS:<br>
H0: pH4j in the ALL group is equal to 0.5<br>
HA: pH4j in the ALL group is less than 0.5<br>
<br>
Since p-value 0.1239 is large, the data did not provide enough evidence to reject
null hypothesis at the significant level 0.05.
```{r}
# Problem 1-d
n.all<-sum(golub[2972,gol.fac=="ALL"]>-0.6)
size.all <-length(golub[2972,gol.fac=="ALL"])
binom.test(n.all,size.all,0.5,alternative="less") # p-value =0.1239
```
<br>
<br>
e.The proportion pH4j in the ALL group differs from the proportion pH4j in the
AML group.<br>
<br>
ANS:<br>
H0: The proportion pH4j in the ALL group is the same from the proportion pH4j in
the AML group<br>
HA: The proportion pH4j in the ALL group differs from the proportion pH4j in the
AML group<br>
<br>
The p-value 0.101 is not significant (less than 0.05), we cannot reject the null
hypothesis. Also, the data does not provide evidence that the proportion pH4j in the
ALL group differs from the proportion pH4j in the AML group.
```{r}
# Problem 1-e
n.aml<-sum(golub[2972,gol.fac=="AML"]>-0.6)
size.aml <-length(golub[2972,gol.fac=="AML"])
prop.test(x=c(n.all,n.aml),n=c(size.all,size.aml),alternative="two.sided") # p-value = 0.101 
```
<br>
<br>
<br>
<br>
**Problem 2 <br>**
Suppose that the probability to reject a biological hypothesis by the
results of a certain experiment is 0.03. This experiment is repeated 3000 times.<br>
a.How many rejections do you expect?<br>
```{r}
reject.p<-0.03
n<-3000
reject.p*n
```
<br>
<br>
b.What is the probability of less than 75 rejections?<br>
```{r}
pbinom(74,size=n,prob=reject.p)
```
<br>
<br>
<br>
<br>
**Problem 3 <br>**
For testing H0: μ=5 versus HA: μ>5, we considers a new α=0.1 level test which
rejects when t_obs=(X̄-5)/(s/√n) falls between t_(0.3,n-1) and t_(0.4,n-1). <br>
<br>
Use a Monte Carlo simulation to estimate the Type I error rate of this test when
n=30. Do 10,000 simulation runs of data sets from the 𝑁(𝜇 = 5, 𝜎 = 4). Please
show the R script for the simulation, and the R outputs for running the script.
Provide your numerical estimate for the Type I error rate. Is this test valid (that is,
is its Type I error rate same as the nominal α=0.1 level)?<br>
<br>
ANS:<br>
The test is valid. The Monte Carlo estimate of the Type I error rate is 0.1020000
with its 95% CI as (0.0960682, 0.1079318). This does agree with the nominal level
of α = 0.1
```{r}
x.sim <- matrix(rnorm(10000*30,mean=5,sd=4),ncol = 30)
tstat<-function(x) (mean(x)-5)/sd(x)*sqrt(length(x))
tstat.sim<-apply(x.sim,1,tstat)
reject.area<-qt(c(0.3,0.4),df=29)
power.sim<-mean(tstat.sim<reject.area[2]&tstat.sim>reject.area[1]) # [1] -0.5302144 -0.2556836
power.sim+c(-1,0,1)*qnorm(0.975)*sqrt(power.sim*(1-power.sim)/10000)
# [1] 0.0960682 0.1020000 0.1079318
```
<br>
<br>
<br>
<br>
**Problem 4 <br>**
On the Golub et al. (1999) data set, do Welch two-sample t-tests to compare
every gene’s expression values in ALL group versus in AML group.<br>
<br>
a.Use Bonferroni and FDR adjustments both at 0.05 level. How many genes
are differentially expressed according to these two criteria?
```{r}
p.values <- apply(golub, 1, function(x) t.test(x)$p.value)
gol.fac <- factor(golub.cl,levels=0:1, labels= c("ALL","AML"))

all.p.values<-p.values[gol.fac=="ALL"]
aml.p.values<-p.values[gol.fac=="AML"]

p.all <-p.adjust(p=all.p.values, method="bonferroni")<0.05 & p.adjust(p=all.p.values, method="fdr")<0.05
sum(p.all) # [1] 1311

p.aml <-p.adjust(p=aml.p.values, method="bonferroni")<0.05 & p.adjust(p=aml.p.values, method="fdr")<0.05
sum(p.aml) # [1] 550
```
<br>
<br>
b.Find the gene names for the top three strongest differentially expressed
genes (i.e., minimum p-values). 
```{r}
# Problem 4-b
golub.gnames[order(all.p.values)[1:3],2]
#[1] "Guanine nucleotide exchange factor p532 mRNA"                   
#[2] "KIAA0094 gene, partial cds"                                     
#[3] "Renal cell carcinoma antigen RAGE-4 mRNA, complete putative cds"

golub.gnames[order(aml.p.values)[1:3],2]
#[1] "PF4 Platelet factor 4"      "IDO Indole 2,3-dioxygenase"
#[3] "APOC1 Apolipoprotein CI" 
```
<br>
<br>
<br>
<br>

## Module 7 Wilcoxon Shapiro-Wilk<br>
**Problem 1 <br>** 
For the Golub et al. (1999) data set, use appropriate Wilcoxon two-sample tests to
find the genes whose mean expression values are higher in the ALL group than in
the AML group. Use FDR adjustments at the 0.05 level. How many genes are
expressed higher in the ALL group? Find the gene names for the top three genes
with smallest p-values.<br>
```{r warning=FALSE}
gol.fac <- factor(golub.cl,levels=0:1, labels= c("ALL","AML"))
ALL.higher = rep(NA, nrow(golub))
for (i in 1:nrow(golub)){
  ALL.higher[i]<-wilcox.test(golub[i,]~gol.fac,paired=F, alternative="two.sided")$p.value
}
p.ALL.higher<- p.adjust(p=ALL.higher, method="fdr")<0.05
sum(p.ALL.higher) # [1] 679
golub.gnames[order(ALL.higher)[1:3],2] # Use Original P-value
```
<br>
<br>
<br>
<br>
**Problem 2 <br>** 
For the Golub et al. (1999) data set, apply the Shapiro-Wilk test of normality to
every gene’s expression values in the AML group. How many genes do not pass
the test at 0.05 level with FDR adjustment? <br>
```{r}
gol.fac <- factor(golub.cl,levels=0:1, labels= c("ALL","AML"))
not.Pass = rep(NA, nrow(golub))
for (i in 1:nrow(golub)){
  not.Pass[i]<-shapiro.test(golub[i,gol.fac == "AML"])$p.value
}
p.not.Pass<- p.adjust(p=not.Pass, method="fdr")>0.05 # not pass
sum(p.not.Pass) #[1] 2826
```
<br>
<br>
<br>
<br>
**Problem 3 <br>** 
Gene "HOXA9 Homeo box A9" can cause leukemia (Golub et al., 1999). Use
appropriate Wilcoxon two-sample tests to test if, for the ALL patients, the gene
"HOXA9 Homeo box A9" expresses at the same level as the “CD33” gene. Please
submit your R script with the answer.<br>
<br>
ANS:<br>
We can see that, for 28 out of 38 pairs, the expression of CD33 gene is bigger than
the expression of the HOXA9 Homeo box A9 gene. This results in a p-value =
0.005098. Hence, the null hypothesis is rejected: the two genes do express
differently.
```{r}
gol.fac <- factor(golub.cl,levels=0:1, labels= c("ALL","AML"))
grep("HOXA9 Homeo box A9",golub.gnames[,2]) # [1] 1391
grep("CD33",golub.gnames[,2]) # [1] 808
diff<-golub[808,gol.fac=="ALL"]-golub[1391,gol.fac=="ALL"]
binom.test(x=sum(diff>0), n=length(diff), p=0.5, alternative="two.sided")
```
<br>
<br>
<br>
<br>


**Problem 4 <br>** 
The data set “UCBAdmissions” in R contains admission decisions by gender at six
departments of UC Berkeley. For this data set, carry out appropriate test for
independence between the admission decision and gender for each of the
departments. What are your conclusions? <br>
<br>
ANS:<br>
H0: The admission decision and gender are independent. (odds ratio equals to one.)<br>
The p-value 2.2e-16 is very small, so we can reject the null hypothesis and
conclude that the admission decision and gender are not independent.
```{r}
testTable<- matrix(c(sum(UCBAdmissions["Admitted","Male",]), sum(UCBAdmissions["Rejected","Male",]), sum(UCBAdmissions["Admitted","Female",]), sum(UCBAdmissions["Rejected","Female",])), nrow=2, dimnames=list("Admit"=c("Admitted","Rejected"), "Gender"=c("Male","Female")))
fisher.test(testTable)
```
<br>
<br>
<br>
<br>
**Problem 5 <br>** 
There are two random samples X1…Xn and Y1…Ym with population means μX and
μY and population variances σ2X and σ2Y.For testing H0: σ2X = σ2Y versus
HA: σ2X < σ2Y, we can use a permutation test for the statistic.<br>
<br>
Please program this permutation test in R. Use this nonparametric test on the
“CD33” gene of the Golub et al. (1999) data set. Test whether the variance in the
ALL group is smaller than the variance in the AML group. <br>
<br>
ANS:<br>
H0: The variance in the ALL and AML group are the same<br>
<br>
Since the p-value of the permutation test is very close to 1, this result suggests the
observed effect almost precisely equals the null hypothesis so that we can conclude
the variance in AML and ALL group are almost the same.
```{r}
gol.fac <- factor(golub.cl,levels=0:1, labels= c("ALL","AML"))
gene.data<- golub[808,]
n<-sum(gol.fac=="ALL")
m<-sum(gol.fac=="AML")
S.obs<-sd(golub[gol.fac=="ALL"])^2/sd(golub[gol.fac=="AML"])^2
n.perm=2000
S.perm = rep(NA, n.perm)
for (i in 1:n.perm){
  data.perm = sample(gene.data, n+m, replace=F)
  S.perm[i] = sd(data.perm[gol.fac=="ALL"])^2/sd(data.perm[gol.fac=="AML"])^2
}
2*min(mean(S.perm<=S.obs), mean(S.perm>=S.obs)) #[1] 0.987
```
